---
title: "Novel Transcripts Workflow"
output: html_document
---

```{r libraries, message=FALSE, warning=FALSE}
library(splicelogic)
library(SummarizedExperiment)
library(GenomicFeatures)
library(dplyr)
library(ggplot2)
library(tibble)
library(here)
library(plyxp)
devtools::load_all(here("../splicelogic"))
```

## Load bambu output

`bambu_discovery.rds` is the exon-by-transcript `GRangesList` from bambu's
discovery run (novel plus annotated transcripts); `bambu_quant_se.rds` is the
matching quantification `SummarizedExperiment`.

```{r setup_bambu, message=FALSE}
bdisc    <- readRDS(here("workflows", "data", "bambu_discovery.rds"))
se_bambu <- readRDS(here("workflows", "data", "bambu_quant_se.rds"))
bdisc
```

## Transcript annotation (condition, CPM, direction)

Condition and donor live only in the sample names; `ctrl` -> WT and `exp` -> KD
to match the satuRn workflow. For each transcript we take the mean CPM in each
group and the `log2(WT / KD)` fold change, whose sign gives the direction.

```{r condition_cpm}
cn <- colnames(se_bambu)
colData(se_bambu)$condition <- factor(
  ifelse(grepl("_ctrl", cn), "WT", "KD"), levels = c("KD", "WT")
)
colData(se_bambu)$donor <- sub("-0008.*$", "", cn)
table(colData(se_bambu)$donor, colData(se_bambu)$condition)

cpm   <- assay(se_bambu, "CPM")
is_wt <- colData(se_bambu)$condition == "WT"

rowData(se_bambu)$cpm_WT <- rowMeans(cpm[,  is_wt, drop = FALSE])
rowData(se_bambu)$cpm_KD <- rowMeans(cpm[, !is_wt, drop = FALSE])
rowData(se_bambu)$lfc_WT_vs_KD <- log2((rowData(se_bambu)$cpm_WT + 1) /
                                       (rowData(se_bambu)$cpm_KD + 1))
rowData(se_bambu)$direction <- ifelse(
  rowData(se_bambu)$cpm_WT > rowData(se_bambu)$cpm_KD, "up_in_WT",
  ifelse(rowData(se_bambu)$cpm_WT < rowData(se_bambu)$cpm_KD, "up_in_KD", "tie")
)

# most `tie` rows are transcripts at zero in both groups
table(rowData(se_bambu)$direction)
table(rowData(se_bambu)$direction[rowMeans(cpm) > 1])
```

A quick look at the novel genes' expression, to see how much signal there is.

```{r explore_novel, message=FALSE}
xp <- se_bambu |> new_plyxp()
xp_novel <- xp |>
  filter(rows(novelGene)) |>
  mutate(rows(mean_cpm = rowMeans(.assays_asis$CPM)))
nrow(xp_novel)

table(cut(log10(rowData(xp_novel)$mean_cpm), -3:3))
hist(log10(rowData(xp_novel)$mean_cpm))
```

### Flat exon GRanges

A flat `GRanges`, one range per exon, with the transcript-level annotation
joined on. This is the exon object the splicelogic `find_*()` and `plot_*()`
functions take.

```{r build_exons}
exons <- unlist(bdisc, use.names = TRUE)
exons$tx_id <- names(exons)

rd <- rowData(se_bambu)
i  <- match(exons$tx_id, rd$TXNAME)
stopifnot(!anyNA(i))

cols    <- c("GENEID", "novelGene", "novelTranscript", "txClassDescription",
             "NDR", "readCount", "relReadCount",
             "cpm_WT", "cpm_KD", "lfc_WT_vs_KD", "direction")
tx_info <- rd[, cols]
tx_info$mean_cpm <- rowMeans(assay(se_bambu, "CPM"))
tx_info$n_det    <- rowSums(assay(se_bambu, "counts") > 0)

mcols(exons) <- cbind(mcols(exons), tx_info[i, ])
names(mcols(exons))[names(mcols(exons)) == "GENEID"] <- "gene_id"

# bambu exons carry no exon_id. Build one from gene + coordinates (per the
# splicelogic vignette), so the same exon gets the same id in every transcript
# that contains it.
exons$exon_id <- paste0(exons$gene_id, ":",
                        start(exons), "-", end(exons))
```

### Novel transcript counts

The comparison keeps novel transcripts at **annotated** loci (so there is an
annotated partner to compare against) and drops novel transcripts at entirely
new loci.

```{r novel_counts}
count_gr <- function(gr, label) {
  data.frame(set   = label,
             exons = length(gr),
             tx    = length(unique(gr$tx_id)),
             genes = length(unique(gr$gene_id)))
}

rbind(
  count_gr(exons, "all (extended annotation)"),
  count_gr(exons[exons$novelTranscript],  "novel transcripts"),
  count_gr(exons[!exons$novelTranscript], "annotated transcripts"),
  count_gr(exons[exons$novelTranscript &  exons$novelGene], "  novel tx, NEW locus (drop)"),
  count_gr(exons[exons$novelTranscript & !exons$novelGene], "  novel tx, annotated gene (keep)")
) |> print(row.names = FALSE)
```

## Splice event detection with splicelogic

The two partitions are defined explicitly (no DTU test): the **up** set is the
novel transcripts at annotated loci, the **down** (reference) set is the
annotated transcripts. `prepare_exons_by_partition()` assigns `estimate = +1`
to up and `-1` to down.

```{r find_events}
up   <- exons |> dplyr::filter(!novelGene, novelTranscript)
down <- exons |> dplyr::filter(!novelGene, !novelTranscript)

# genes present in both partitions are the ones with a comparison to make
length(intersect(unique(up$gene_id), unique(down$gene_id)))

sl_exons <- prepare_exons_by_partition(up, down) |>
  preprocess(coef_col = "estimate")

events_bambu <- find_all_events(sl_exons)
```

An exon paired with several partner transcripts is reported once per partner,
as repeated rows with the same range. Collapsing those per event type shows how
much of the raw event count is that redundancy.

```{r collapse_events}
# define a helper function
my_paste <- \(x) paste(unique(x), collapse = ",")

collapsed_events <- events_bambu |>
  dplyr::group_by(exon_id, event_type) |>
  plyranges::reduce_ranges_directed(
    n           = plyranges::n(),
    tx_id       = my_paste(tx_id),
    exon_rank   = my_paste(exon_rank),
    event_tx_id = my_paste(event_tx_id)
  )
```

```{r barplot_event_types}
event_counts <- events_bambu |>
  as_tibble() |>
  dplyr::count(event_type, name = "count")

collapsed_event_counts <- collapsed_events |>
  as_tibble() |>
  dplyr::count(event_type, name = "count")

# combine both count tables into one, labelled by source
counts <- dplyr::bind_rows(
  "All events" = event_counts,
  "Collapsed events" = collapsed_event_counts,
  .id = "set"
)

ggplot(counts, aes(x = event_type, y = count, fill = set)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c(
    "All events" = "#9ecae1",       # light blue
    "Collapsed events" = "#3182bd"  # darker blue
  )) +
  labs(title = "Novel-transcript splice event counts",
       x = "Event Type", y = "Count", fill = NULL) +
  theme_bw()
```

## Per-transcript annotation for the plots

One label per transcript (novel status, then the direction and the two group
means) and a matching `log2(WT / KD)` value used to draw the LFC strip beside
each transcript.

```{r plot_annotation}
# Pulled into a plain data.frame and read with with(): repeating `rd$` inside
# one call makes the console's `$` completion eat the paste.
d <- as.data.frame(rd[, c("TXNAME", "novelTranscript",
                          "cpm_WT", "cpm_KD", "lfc_WT_vs_KD")])

# novel/annot status, plus the direction and group means. A direction on two
# near-zero means is rounding noise, not a change, so those read "n.d.".
tx_annot <- with(d, paste0(
  ifelse(novelTranscript, "novel", "annot"), " | ",
  ifelse(cpm_WT + cpm_KD < 0.1, "n.d.",
         sprintf("log2 %+.2f  (WT %.2f / KD %.2f)", lfc_WT_vs_KD, cpm_WT, cpm_KD))
))
names(tx_annot) <- d[["TXNAME"]]

# log2(WT/KD) for the LFC strip: NA (no bar) where both groups are near zero.
tx_lfc <- with(d, ifelse(cpm_WT + cpm_KD < 0.1, NA_real_, lfc_WT_vs_KD))
names(tx_lfc) <- d[["TXNAME"]]

# one shared scale so bar lengths mean the same thing on every page
lfc_max <- quantile(abs(tx_lfc), 0.99, na.rm = TRUE)
```

## Plotting helpers

Two small helpers that the package does not provide: a multi-page PDF writer,
and a per-gene layout that places `plot_transcripts()` next to the companion
`plot_transcript_lfc()` strip (the package draws them as separate plots; here
`patchwork` composes them so they share row order and y limits).

```{r plot_helpers}
fig_dir <- here("workflows", "figures")
dir.create(fig_dir, showWarnings = FALSE)

# write a list of plots to one multi-page PDF, one plot per page
write_pdf <- function(plots, file, width = 10, height = 6) {
  grDevices::pdf(file, width = width, height = height, onefile = TRUE)
  on.exit(grDevices::dev.off(), add = TRUE)
  for (nm in names(plots)) print(plots[[nm]])
  invisible(file)
}

# plot_transcripts() + plot_transcript_lfc() side by side, one pair per gene
plot_gene_with_lfc <- function(events, exons, tx_lfc, lfc_max = NULL,
                               tx_annot = NULL, lfc_ends = c("KD", "WT"),
                               widths = c(5, 1.2), ...) {
  genes <- unique(as.character(events$gene_id))
  plots <- lapply(stats::setNames(genes, genes), function(g) {
    ev <- events[as.character(events$gene_id) == g]
    tryCatch({
      p <- plot_transcripts(ev, exons, title = g, tx_annot = tx_annot, ...)
      q <- plot_transcript_lfc(ev, tx_lfc, lfc_max = lfc_max,
                               lfc_ends = lfc_ends)
      patchwork::wrap_plots(p, q, widths = widths)
    }, error = function(e) {
      warning("Skipping ", g, ": ", conditionMessage(e))
      NULL
    })
  })
  Filter(Negate(is.null), plots)
}
```

## Events for one gene

`exons` is already the flat `GRanges` the package plotting functions take (one
range per exon with a `tx_id` column), so it is passed directly.

```{r plot_one_gene, fig.width=10, fig.height=6}
gene_of_interest <- unique(as.character(events_bambu$gene_id))[1]
events_for_gene  <- events_bambu[events_bambu$gene_id == gene_of_interest]

p <- plot_transcripts(events_for_gene, exons, title = gene_of_interest,
                      tx_annot = tx_annot)
q <- plot_transcript_lfc(events_for_gene, tx_lfc, lfc_max = lfc_max,
                         lfc_ends = c("KD", "WT"))
patchwork::wrap_plots(p, q, widths = c(5, 1.2))
```

## All events, one page per gene (with the LFC strip)

```{r plot_by_gene, eval=FALSE}
gene_plots <- plot_gene_with_lfc(
  events_bambu, exons,
  tx_annot = tx_annot, tx_lfc = tx_lfc, lfc_max = lfc_max
)
length(gene_plots)

write_pdf(gene_plots, file.path(fig_dir, "events_by_gene_bambu.pdf"))
```

## All events, one page per transcript pair

`plot_transcripts_by_pair()` splits on the transcript pair internally, so
`events_bambu` works directly.

```{r plot_by_pair, eval=FALSE}
pair_plots <- plot_transcripts_by_pair(events_bambu, exons)
length(pair_plots)

write_pdf(pair_plots, file.path(fig_dir, "events_by_pair_bambu.pdf"), height = 3)
```

## Novel transcripts split by direction

Plot the novel transcripts separately by the direction of their WT/KD change.
In each event row the novel transcript is the one from the `up` partition
(`estimate > 0`): `tx_id` when the row's own estimate is positive, otherwise its
partner `event_tx_id`.

```{r novel_by_direction}
ev <- as.data.frame(S4Vectors::mcols(events_bambu)[, c("tx_id", "event_tx_id",
                                                       "estimate", "gene_id")])
novel_tx  <- with(ev, ifelse(estimate > 0, as.character(tx_id),
                             as.character(event_tx_id)))
novel_lfc <- unname(tx_lfc[novel_tx])   # NA when the novel transcript is n.d.

# which() drops the NA rows, so n.d. novel transcripts are in neither set
events_novel_kd <- events_bambu[which(novel_lfc < 0)]
events_novel_wt <- events_bambu[which(novel_lfc > 0)]

n_genes <- function(keep) length(unique(ev[["gene_id"]][keep]))
data.frame(
  set    = c("all events", "novel tx higher in KD (< 0)",
             "novel tx higher in WT (> 0)", "novel tx n.d. (not plotted)"),
  events = c(nrow(ev), sum(novel_lfc < 0, na.rm = TRUE),
             sum(novel_lfc > 0, na.rm = TRUE), sum(is.na(novel_lfc))),
  genes  = c(n_genes(TRUE), n_genes(which(novel_lfc < 0)),
             n_genes(which(novel_lfc > 0)), n_genes(is.na(novel_lfc)))
)

novel_sets <- list(novel_higher_in_KD = events_novel_kd,
                   novel_higher_in_WT = events_novel_wt)
```

A gene with novel transcripts in both directions appears in both PDFs, each
page showing only the novel transcripts of that direction and their partners.

```{r plot_novel_by_direction, eval=FALSE}
for (set in names(novel_sets)) {
  set_plots <- plot_gene_with_lfc(
    novel_sets[[set]], exons,
    tx_annot = tx_annot, tx_lfc = tx_lfc, lfc_max = lfc_max
  )
  set_file <- file.path(fig_dir, paste0("events_", set, "_bambu.pdf"))
  write_pdf(set_plots, set_file)
  message(set, ": ", length(set_plots), " pages written to ", set_file)
}
```

## Splice-site logos for the novel-transcript events

Every event exon is a real spliced exon, so its flanking splice sites should be
the canonical ones: the **3' acceptor** (`...AG |` exon) in the region
**upstream** of the exon, and the **5' donor** (exon `| GU...`) **downstream**.
This is a sanity check that the novel-transcript events sit at genuine splice
sites. 

```{r logo_helpers, message=FALSE}
source(here("workflows/helper.R"))
```

```{r novel_splice_logo, message=FALSE, warning=FALSE, fig.width=9, fig.height=3.2}
splice_events <- events_bambu[!events_bambu$event_type %in% c("atss", "ates")]

# Restrict to the standard chromosomes: the bambu object carries hundreds of
# _fix/_alt scaffolds that are not all in the BSgenome, which otherwise triggers
# a noisy seqinfo-merge warning (and errors for exons on absent scaffolds).
# pruning.mode = "coarse" drops ranges on non-standard seqlevels.
splice_events <- GenomeInfoDb::keepStandardChromosomes(
  splice_events, pruning.mode = "coarse"
)
GenomeInfoDb::seqlevels(splice_events) <-
  GenomeInfoDb::seqlevelsInUse(splice_events)

# 10 bp into the intron on each side (exon_bases = 0, so purely intronic):
# the upstream logo should end in AG, the downstream logo should start with GU.
novel_logo <- slFlankLogoMatrices(splice_events, genome = "hg38",
                                  width = 10, exon_bases = 0)

slPlotFlankLogo(novel_logo, method = "prob")
```

### Non-canonical splice sites

To find the events that break the pattern, take the 2 intronic bases
immediately flanking each exon — strand-aware, via `plyranges::flank_*()` — and
read them with `get_seq()`. The acceptor (upstream) should be `AG`, the donor
(downstream) `GU`. Anything else is flagged.

```{r splice_dinucleotides, message=FALSE}
# one row per distinct exon, so an exon paired with several partners is not
# counted repeatedly
ex_once <- slUniqueExons(splice_events)

# 2 bp immediately 5' (acceptor) and 3' (donor) of each exon, in transcription
# order; get_seq() reverse-complements minus-strand ranges automatically
acceptor <- ex_once |> plyranges::flank_upstream(width = 2)
donor    <- ex_once |> plyranges::flank_downstream(width = 2)

ex_once$acceptor <- as.character(get_seq(acceptor, "hg38", as_rna = TRUE))
ex_once$donor    <- as.character(get_seq(donor,    "hg38", as_rna = TRUE))
ex_once$canonical <- ex_once$acceptor == "AG" & ex_once$donor == "GU"

# how many follow the canonical AG/GU pattern
table(canonical = ex_once$canonical)
table(acceptor = ex_once$acceptor)
table(donor = ex_once$donor)
```

```{r noncanonical_list}
# the offenders: event exons whose acceptor is not AG or donor is not GU
noncanonical <- ex_once[!ex_once$canonical]

noncanonical |>
  as_tibble() |>
  dplyr::select(seqnames, start, end, strand, gene_id, tx_id, exon_rank,
                event_type, acceptor, donor) |>
  dplyr::arrange(acceptor, donor) |>
  print(n = 20, width = Inf)
```

To eyeball one, read the whole exon with 5 intronic bases on each side, in
transcription order (`get_seq()` reverse-complements the minus strand). The
acceptor is the 2 intronic bases just before the exon (canonical `AG`), the
donor the 2 intronic bases just after it (canonical `GU`):

```{r inspect_one_noncanonical}
one <- noncanonical[1]

up5   <- get_seq(plyranges::flank_upstream(one,   width = 5), "hg38", as_rna = TRUE)
exon  <- get_seq(one,                                         "hg38", as_rna = TRUE)
down5 <- get_seq(plyranges::flank_downstream(one, width = 5), "hg38", as_rna = TRUE)

# intron(5) | EXON | intron(5), 5'->3'
cat(sprintf("%s | %s | %s\n", up5, exon, down5))

data.frame(
  intron_up_5   = as.character(up5),   # ...ends in AG for a canonical acceptor
  exon          = as.character(exon),
  intron_down_5 = as.character(down5)  # starts with GU for a canonical donor
)
```

