---
title: "PCBP1 workflow"
output: html_document
---

Data: "Iron-sensitive RNA regulation by poly C binding proteins" (PMID 41036621),
two GEO series of K562 short-read RNA-seq (polyA, paired-end 150), three
biological replicates per condition, quantified with Salmon (v1.10.2) against
GENCODE v43:

- [GSE279951](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE279951)
  [P1KD]: shRNA against luciferase (`shCtrl`) or PCBP1 (`shPCBP1`).
- [GSE279949](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE279949)
  [chelator]: DMSO (control) or iron chelator 21H7 (`CHEL`), overnight.

```{r libraries, message=FALSE, warning=FALSE}
library(splicelogic)
library(SummarizedExperiment)
library(GenomicFeatures)
library(dplyr)
library(ggplot2)
library(satuRn)
library(limma)
library(tibble)
library(tximeta)
library(here)
devtools::load_all(here("../splicelogic"))
```

## Download data

GEO bundles one Salmon `quant.sf` per sample into a `RAW.tar` per series
(~21MB each). They are extracted into `workflows/data/PCBP1/<GSE>/`.

GEO has only the `quant.sf` files, but tximeta also needs the Salmon
`aux_info/meta_info.json` (it identifies the transcriptome from the index hash
stored there). The authors' `meta_info.json` is not on GEO: place it at
`workflows/data/PCBP1/meta_info.json` and it is copied into an `aux_info/`
folder next to the quant files of each series (same Salmon index for both).

```{r download_data, eval=FALSE}
options(timeout = 600)
geo_dir <- here("workflows/data/PCBP1")
for (gse in c("GSE279949", "GSE279951")) {
  dest <- file.path(geo_dir, gse)
  dir.create(dest, recursive = TRUE, showWarnings = FALSE)
  tar_file <- file.path(dest, paste0(gse, "_RAW.tar"))
  download.file(
    paste0("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE279nnn/", gse,
           "/suppl/", gse, "_RAW.tar"),
    tar_file, mode = "wb"
  )
  untar(tar_file, exdir = dest)
  file.remove(tar_file)
  dir.create(file.path(dest, "aux_info"), showWarnings = FALSE)
  file.copy(file.path(geo_dir, "meta_info.json"), file.path(dest, "aux_info"))
}
list.files(geo_dir, recursive = TRUE)
```

```{r txdb, eval=FALSE}
library(txdbmaker)
# load TxDb for GRanges conversion
# the quant.sf files use GENCODE v43 transcript IDs (with version), so build the
# TxDb from the same release: the v48 TxDb would drop transcripts whose version changed
# download once (warning: ~50MB compressed, slow from EBI)
gtf <- here("workflows/data/gencode.v43.annotation.gtf.gz")
# curl with -C - resumes a partial file (EBI often times out mid-transfer);
# --retry re-attempts on failure instead of restarting from zero
system2("curl", c("-C", "-", "-L", "--retry", "5", "--retry-delay", "5",
                  "-o", shQuote(gtf),
                  "https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_43/gencode.v43.annotation.gtf.gz"))

txdb <- txdbmaker::makeTxDbFromGFF(gtf, format = "gtf")
# GENCODE v43 has ~252k transcripts; far fewer means the GTF download was cut short
length(AnnotationDbi::keys(txdb, "TXNAME"))
# save to not rebuild every time
AnnotationDbi::saveDb(txdb, here("workflows/data/gencode.v43.sqlite"))
```

## Build the SummarizedExperiment

One `SummarizedExperiment` with the 12 samples of both series, imported with
tximeta. The sample table is parsed from the file names,
`<GSM>_<group>_<replicate>_quant.sf.txt.gz`.

```{r make_se, eval=FALSE}
txdb <- AnnotationDbi::loadDb(here("workflows/data/gencode.v43.sqlite"))
tx2gene <- AnnotationDbi::select(txdb, keys(txdb, "TXNAME"), "GENEID", "TXNAME")

coldata <- data.frame(
  files = list.files(here("workflows/data/PCBP1"),
                     pattern = "_quant\\.sf\\.txt\\.gz$",
                     recursive = TRUE, full.names = TRUE)
) |>
  dplyr::mutate(
    series = basename(dirname(files)),
    geo_accession = sub("_.*", "", basename(files)),
    group = sub("^GSM[0-9]+_(.+)_[0-9]+_quant.*", "\\1", basename(files)),
    replicate = sub(".*_([0-9]+)_quant.*", "\\1", basename(files)),
    names = paste(group, replicate, sep = "_"),
    experiment = dplyr::case_when(
      series == "GSE279951" ~ "PCBP1_KD",
      series == "GSE279949" ~ "chelator"
    ),
    condition = dplyr::case_when(
      group == "shCtrl" ~ "WT",
      group == "shPCBP1" ~ "KD",
      group == "DMSO" ~ "DMSO",
      group == "CHEL" ~ "CHEL"
    )
  )
coldata

# check how many transcripts in the quant files are in txdb (should be 1;
# tximeta below fails if any are missing)
read.delim(coldata$files[1])$Name %in% tx2gene$TXNAME |> mean()

# tximeta matches the index hash in aux_info/meta_info.json to GENCODE v43, then
# looks in its BiocFileCache for a TxDb named after the v43 GTF: register the
# TxDb built above under that name, so tximeta loads it instead of downloading
# and building its own
bfc <- BiocFileCache::BiocFileCache(tximeta:::getBFCLoc(), ask = FALSE)
in_cache <- BiocFileCache::bfcquery(
  bfc, "gencode.v43.annotation.gtf.gz", field = "rname", exact = TRUE
)
if (nrow(in_cache) == 0) {
  BiocFileCache::bfcadd(
    bfc, rname = "gencode.v43.annotation.gtf.gz",
    fpath = here("workflows/data/gencode.v43.sqlite"), action = "asis"
  )
}

# transcript-level counts; dtuScaledTPM scales TPM by the median transcript
# length of each gene, the tximport setting recommended for DTU
se <- tximeta::tximeta(
  coldata, countsFromAbundance = "dtuScaledTPM", tx2gene = tx2gene
)

se

saveRDS(se, here("workflows/data/PCBP1/PCBP1_SummarizedExperiment.rds"))
```

```{r setup, message=FALSE}
se <- readRDS(here("workflows/data/PCBP1/PCBP1_SummarizedExperiment.rds"))
dim(se)
colData(se)
```

## DTU testing with satuRn: PCBP1 knockdown (GSE279951)

We compare control shRNA (`WT`) against PCBP1 knockdown (`KD`) using satuRn.
GEO gives no pairing between the replicates of the two conditions, so there is
no blocking term (unlike `donor` in the RBP workflow).

```{r filter_kd}
se_kd <- se[, se$experiment == "PCBP1_KD"]
dim(se_kd)

# counts >= 10 in at least 3 samples (the size of one group)
keep <- rowSums(assay(se_kd) >= 10) >= 3
se_kd <- se_kd[keep, ]
dim(se_kd)
```

```{r dtu_design_kd}
# KD as reference level
colData(se_kd)$condition <- factor(colData(se_kd)$condition, levels = c("KD", "WT"))

# build design from the factor directly (matches satuRn vignette convention)
condition <- colData(se_kd)$condition
design <- model.matrix(~ 0 + condition)

L <- limma::makeContrasts(
  WT_vs_KD = conditionWT - conditionKD,
  levels = design
)
L
```

```{r dtu_fit_kd, message=FALSE}
rowData(se_kd)$isoform_id <- rowData(se_kd)$tx_name
# flatten CharacterList gene_id to plain character
rowData(se_kd)$gene_id <- vapply(rowData(se_kd)$gene_id, `[`, character(1), 1)

se_kd <- se_kd[
  duplicated(rowData(se_kd)$gene_id) |
    duplicated(rowData(se_kd)$gene_id, fromLast = TRUE),
]
dim(se_kd)
se_kd <- satuRn::fitDTU(se_kd, formula = ~0 + condition)
```

```{r dtu_test_kd, message=FALSE}
se_kd <- satuRn::testDTU(
  object = se_kd,
  contrasts = L,
  diagplot1 = TRUE,
  diagplot2 = TRUE
)
```

```{r dtu_results_kd}
gene_ids <- rowData(se_kd)$gene_id

# every tested transcript; `sig_res_kd` below is the significant subset, and the
# non-significant rows are the background pool for slSampleNonEventExons()
dtu_res_kd <- rowData(se_kd)[["fitDTUResult_WT_vs_KD"]] |>
  as.data.frame() |>
  tibble::rownames_to_column("tx_name") |>
  tibble::as_tibble() |>
  dplyr::mutate(gene_id = gene_ids) |>
  dplyr::select(tx_name, gene_id, estimates, pval, regular_FDR) |>
  dplyr::arrange(pval)

sig_res_kd <- dtu_res_kd |> dplyr::filter(regular_FDR < 0.1)

sig_res_kd <- sig_res_kd |> mutate(tx_id = tx_name)
sig_res_kd
# save for later use
saveRDS(dtu_res_kd, here("workflows/data/PCBP1/dtu_res_kd.rds"))
saveRDS(sig_res_kd, here("workflows/data/PCBP1/sig_res_kd.rds"))
```



## DTU testing with satuRn: iron chelation (GSE279949)

Same steps, comparing DMSO (`DMSO`) against the iron chelator (`CHEL`).

```{r filter_chel}
se_chel <- se[, se$experiment == "chelator"]
dim(se_chel)

# counts >= 10 in at least 3 samples (the size of one group)
keep <- rowSums(assay(se_chel) >= 10) >= 3
se_chel <- se_chel[keep, ]
dim(se_chel)
```

```{r dtu_design_chel}
# CHEL as reference level
colData(se_chel)$condition <- factor(colData(se_chel)$condition, levels = c("CHEL", "DMSO"))

# build design from the factor directly (matches satuRn vignette convention)
condition <- colData(se_chel)$condition
design <- model.matrix(~ 0 + condition)

L <- limma::makeContrasts(
  DMSO_vs_CHEL = conditionDMSO - conditionCHEL,
  levels = design
)
L
```

```{r dtu_fit_chel, message=FALSE}
rowData(se_chel)$isoform_id <- rowData(se_chel)$tx_name
# flatten CharacterList gene_id to plain character
rowData(se_chel)$gene_id <- vapply(rowData(se_chel)$gene_id, `[`, character(1), 1)

se_chel <- se_chel[
  duplicated(rowData(se_chel)$gene_id) |
    duplicated(rowData(se_chel)$gene_id, fromLast = TRUE),
]
dim(se_chel)
se_chel <- satuRn::fitDTU(se_chel, formula = ~0 + condition)
```

```{r dtu_test_chel, message=FALSE}
se_chel <- satuRn::testDTU(
  object = se_chel,
  contrasts = L,
  diagplot1 = TRUE,
  diagplot2 = TRUE
)
```

```{r dtu_results_chel}
gene_ids <- rowData(se_chel)$gene_id

dtu_res_chel <- rowData(se_chel)[["fitDTUResult_DMSO_vs_CHEL"]] |>
  as.data.frame() |>
  tibble::rownames_to_column("tx_name") |>
  tibble::as_tibble() |>
  dplyr::mutate(gene_id = gene_ids) |>
  dplyr::select(tx_name, gene_id, estimates, pval, regular_FDR) |>
  dplyr::arrange(pval)

sig_res_chel <- dtu_res_chel |> dplyr::filter(regular_FDR < 0.3)

sig_res_chel <- sig_res_chel |> mutate(tx_id = tx_name)
sig_res_chel
# save for later use
saveRDS(dtu_res_chel, here("workflows/data/PCBP1/dtu_res_chel.rds"))
saveRDS(sig_res_chel, here("workflows/data/PCBP1/sig_res_chel.rds"))
```

```{r load_sig_res}
# load the saved DTU results
sig_res_kd <- readRDS(here("workflows/data/PCBP1/sig_res_kd.rds"))
exsig_res_chel <- readRDS(here("workflows/data/PCBP1/sig_res_chel.rds"))
sig_res_kd
sig_res_chel
```

## Splice event detection with splicelogic for PCBP1 knockdown

We run event detection on the PCBP1 knockdown comparison (`sig_res_kd`); the same
steps apply to the chelator comparison by swapping in `sig_res_chel`.

### Direction of the events

WT (PCBP1 present) is the up side and KD (no PCBP1) the reference. The contrast
is `WT_vs_KD = conditionWT - conditionKD`, so a positive estimate means the
transcript is used more in WT. splicelogic treats the down transcripts as the
reference, so every event describes what WT transcripts do compared with KD
ones, as in the RBP workflow. With `~ 0 + condition` there is no intercept, so
the level order (`levels = c("KD", "WT")`) does not change the fit: the
contrast alone sets the direction.

| Event | What happens | What it suggests about PCBP1 |
|---|---|---|
| IE | Exon is in WT transcripts and lost after KD | PCBP1 promotes inclusion |
| SE | Exon is skipped in WT transcripts and gained after KD | PCBP1 represses inclusion (promotes skipping) |

PCBP1 binding upstream of an exon typically promotes its inclusion, and
knockdown lowers it, so that pattern shows up as IE events.
SE events are kept as the context-dependent group, and the two sets are
compared in the sequence-context analysis below.

The source paper ("Iron-sensitive RNA regulation by poly C binding proteins",
PMID 41036621) reports the same direction with rMATS: after PCBP1 knockdown,
4177 cassette exons were skipped and 2218 included (Fig. 2K), and PCBP1
overexpression shifted the balance the other way (1233 skipped, 2126 included;
Fig. 2M), so PCBP1 primarily promotes exon inclusion. They read the direction
from the KD side and we read it from the WT side, so their exons skipped after
KD are our IE events and their exons included after KD are our SE events. If
our results agree, IE events should outnumber SE events.

```{r make_exons, eval=FALSE}
# build exon GRangesList and rename from internal TXID to TXNAME
txdb <- AnnotationDbi::loadDb(here("workflows/data/gencode.v43.sqlite"))
exons <- exonsBy(txdb, by = "tx")
txps <- AnnotationDbi::select(txdb, keys(txdb, "TXID"), c("TXNAME", "GENEID"), "TXID") |>
  tibble::as_tibble() |>
  dplyr::mutate(TXID = as.character(TXID))
# rename exons to match tx_name in se, matching by TXID
all.equal(names(exons), txps$TXID)
names(exons) <- txps$TXNAME[match(names(exons), txps$TXID)]
saveRDS(exons, here("workflows/data/PCBP1/gencode.v43.exons.rds"))
```

```{r input_formatting}
# load from saved sqlite
txdb <- AnnotationDbi::loadDb(here("workflows/data/gencode.v43.sqlite"))
# load exons from saved rds
exons <- readRDS(here("workflows/data/PCBP1/gencode.v43.exons.rds"))
```

```{r preprare_exons}
# prepare exons for splicelogic
sig_exons <- prepare_exons(
  txdb, sig_res_kd, coef_col = "estimates", verbose = TRUE
)
```

```{r find_events}
#find se
se_events <- sig_exons |> preprocess(coef_col = "estimates")|> find_se()

#find all events
events <- sig_exons |> preprocess(coef_col = "estimates")|> find_all_events()
```

```{r collapse_redundant_events}
# define a helper function
my_paste <- \(x) paste(unique(x), collapse = ",")
collapsed_events <- events |>
    dplyr::group_by(exon_id, event_type) |>
    plyranges::reduce_ranges_directed(
        n = plyranges::n(),
        tx_id = my_paste(tx_id),
        exon_rank = my_paste(exon_rank),
        event_tx_id = my_paste(event_tx_id),
  )
```

```{r barplot_event_types}
event_counts <- events |>
    as_tibble() |>
    dplyr::group_by(event_type) |>
    dplyr::summarise(count = dplyr::n())

collapsed_event_counts <- collapsed_events |>
    as_tibble() |>
    dplyr::group_by(event_type) |>
    dplyr::summarise(count = dplyr::n())

# combine both count tables into one, labelled by source
counts <- dplyr::bind_rows(
  "All events" = event_counts,
  "Collapsed events" = collapsed_event_counts,
  .id = "set"
)

ggplot(counts, aes(x = event_type, y = count, fill = set)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c(
    "All events" = "#9ecae1",       # light blue
    "Collapsed events" = "#3182bd"  # darker blue
  )) +
  labs(title = "Splice event counts",
  x = "Event Type", y = "Count", fill = NULL) +
  theme_bw()
```

```{r plot_events_by_gene}
# use splicelogic's built-in plotting function to visualize events for a specific gene
gene_of_interest <- events$gene_id[1]
# filter events for the gene of interest
events_for_gene <- events[events$gene_id == gene_of_interest]

# visualize the events for the gene of interest
splicelogic::plot_transcripts(events_for_gene, sig_exons, rescale_introns = FALSE) +
  ggtitle(paste("Splice events for gene:", gene_of_interest))
```

```{r save_events_plots, eval=FALSE}
# make a plot for each gene with events and save to files
# plot_transcripts_by_gene returns a list of ggplot objects, one for each gene
plots <- plot_transcripts_by_gene(events, sig_exons)
# save each plot to the same pdf file
pdf(here("workflows/figures/PCBP1_splice_events_by_gene.pdf"), width = 8, height = 6)
for (gene in names(plots)) {
    print(plots[[gene]])
}
dev.off()
```

## Sequence context around the event exons, focusing on included exons (IE)

IE exons are the main set here (see *Direction of the events* above); SE exons
are kept below as a comparison.

The analysis below uses a set of sequence-context helpers consolidated in
`helper.R` (nucleotide-composition profiles, background exon sampling,
splice-site logos and RBP motif scanning). The splicelogic package itself only
exports `get_seq()` for sequence work, so these stay as paper-repo helpers (for now).

```{r seq_context_helpers, message=FALSE}
source(here("workflows/helper.R"))
```

`slEventFlankProfiles()` takes a finder's output straight away: it flanks each
exon (strand-aware, so "upstream" is 5' in transcription order), slides
overlapping windows across the flank, and returns the per-window nucleotide
fractions for each side.

```{r flank_windows_ie, message=FALSE}
ie_events <- sig_exons |> preprocess(coef_col = "estimates") |> find_ie()

ie_profiles <- slEventFlankProfiles(ie_events, genome = "hg38", width = 100)

# rows = events, cols = window_nucleotide (e.g. w1_A)
dim(ie_profiles$upstream)
dim(ie_profiles$downstream)
```

```{r plot_updownstream_ie, fig.width=8, fig.height=3.5}
slPlotFlankProfiles(ie_profiles, exon_label = "IE")
```

### skipped exons events

```{r flank_windows_se, message=FALSE}
se_profiles <- slEventFlankProfiles(se_events, genome = "hg38", width = 100)

# rows = events, cols = window_nucleotide (e.g. w1_A)
dim(se_profiles$upstream)
dim(se_profiles$downstream)
```

```{r plot_updownstream_se, fig.width=8, fig.height=3.5}
slPlotFlankProfiles(se_profiles, exon_label = "SE")
```


## Background set

We want to compare the sequence context of the events with that of a control set.
For the control set (non-event exons), we sample as many exons as there are events,
drawn at random from the internal exons of transcripts with no evidence of
differential usage (`pval > 0.5`).

```{r non_event_exons}
dtu_res_kd <- readRDS(here("workflows/data/PCBP1/dtu_res_kd.rds"))
non_event <- slSampleNonEventExons(
  dtu_res_kd, exons, n = ie_events, pval_min = 0.5, seed = 1
)
non_event
```

```{r flank_windows_non_event, message=FALSE, fig.width=8, fig.height=3.5}
non_event_profiles <- slEventFlankProfiles(
  non_event, genome = "hg38", width = 100
)
dim(non_event_profiles$upstream)

slPlotFlankProfiles(non_event_profiles, exon_label = "non-IE")
```

An alternative, within-transcript control: for each event, the internal exon of
the same transcript that is shared with the event partner and closest in rank
to the event exon.

```{r matched_non_event_exons, eval=FALSE, fig.width=8, fig.height=3.5}
matched <- slMatchedNonEventExons(ie_events, exons)

slEventFlankProfiles(matched, genome = "hg38", width = 100) |>
  slPlotFlankProfiles(exon_label = "matched exon")
```

## Splice-site sequence logos

The window profiles above average over 10 bp, which smooths the splice site
away. `slFlankLogoMatrices()` keeps every base position separate and returns a
position frequency matrix per side, running `width` bases into the intron and
`exon_bases` into the exon so the boundary is inside the matrix.

Note which site each side carries: the region **upstream** of an exon is the
end of the preceding intron and holds the **3' splice site** (acceptor,
`...ag | EXON`); the region **downstream** is the start of the following intron
and holds the **5' splice site** (donor, `EXON | gu...`).

```{r logo_events, message=FALSE, warning=FALSE, fig.width=13, fig.height=3.2}
ie_logo <- slFlankLogoMatrices(ie_events, genome = "hg38",
                               width = 25, exon_bases = 3)

# letter height = frequency at that position
slPlotFlankLogo(ie_logo, method = "prob")
```

```{r logo_events_bits, message=FALSE, warning=FALSE, fig.width=13, fig.height=3.2}
# letter height = information content, the classic logo scaling: unconstrained
# positions shrink to nothing and the consensus stands out
slPlotFlankLogo(ie_logo, method = "bits")
```

The same for the background set, to see whether the event splice sites differ
from those of exons in transcripts with no differential usage.

```{r logo_non_event, message=FALSE, warning=FALSE, fig.width=13, fig.height=3.2}
non_event_logo <- slFlankLogoMatrices(non_event, genome = "hg38",
                                      width = 25, exon_bases = 3)

slPlotFlankLogo(non_event_logo, method = "bits")
```

```{r logo_difference}
# where the two sets differ most, per position and nucleotide; upstream holds
# the 3' splice site, where a C-rich tract preceding the exon would sit
delta_up <- ie_logo$upstream - non_event_logo$upstream
round(delta_up[, order(-apply(abs(delta_up), 2, max))[1:6]], 3)

delta_down <- ie_logo$downstream - non_event_logo$downstream
round(delta_down[, order(-apply(abs(delta_down), 2, max))[1:6]], 3)
```
