---
title: "RBP workflow"
output: html_document
---

```{r libraries, message=FALSE, warning=FALSE}
library(splicelogic)
library(SummarizedExperiment)
library(GenomicFeatures)
library(dplyr)
library(ggplot2)
library(satuRn)
library(limma)
library(tibble)
library(here)
devtools::load_all(here("../splicelogic"))
```

```{r setup, message=FALSE}
# dir <- system.file("extdata", package = "splicelogic")
# se <- readRDS(file.path(dir, "GTEx_v9_ONT_SummarizedExperiment.rds"))
se <- readRDS(here("workflows/data/GTEx_v9_ONT_SummarizedExperiment.rds"))
dim(se)

keep <- rowSums(assay(se) >= 10) >= 6
se <- se[keep, ]
dim(se)
# filters: 385669 -> 18985 transcripts
```

```{r cpm}
library(DESeq2)
library(tximeta)
colsums <- colSums(assay(se, "counts"))
cpm <- t(t(assay(se, "counts")) / colsums * 1e6)
assay(se, "abundance") <- cpm
# dummy lengths... no correction for length
assay(se, "length", withDimnames=FALSE) <- matrix(1000, nrow=nrow(se), ncol=ncol(se))
```

## DTU testing with satuRn

We compare the experiment condition (`exp`, RBP knockdown) against
control (`ctrl`) using satuRn.

```{r dtu_design}
colData(se)$condition <- dplyr::case_when(
    colData(se)$condition == "ctrl" ~ "WT",
    colData(se)$condition == "exp" ~ "KD"
)
# KD as reference level
colData(se)$condition <- factor(colData(se)$condition, levels = c("KD", "WT"))

# build design from the factor directly (matches satuRn vignette convention)
condition <- colData(se)$condition
donor <- factor(colData(se)$donor)
design <- model.matrix(~ 0 + condition + donor)

L <- limma::makeContrasts(
  WT_vs_KD = conditionWT - conditionKD,
  levels = design
)
L
```

```{r dtu_fit, message=FALSE}
rowData(se)$isoform_id <- rowData(se)$tx_name
# flatten CharacterList gene_id to plain character
rowData(se)$gene_id <- vapply(rowData(se)$gene_id, `[`, character(1), 1)

se <- se[
  duplicated(rowData(se)$gene_id) |
    duplicated(rowData(se)$gene_id, fromLast = TRUE),
]
# 18985 transcripts -> 11874 transcripts after removing single-isoform genes
se <- satuRn::fitDTU(se, formula = ~0 + condition + donor)
```

```{r dtu_test, message=FALSE}
se <- satuRn::testDTU(
  object = se,
  contrasts = L,
  diagplot1 = TRUE,
  diagplot2 = TRUE
)
```

```{r dtu_results}
gene_ids <- rowData(se)$gene_id

# every tested transcript; `sig_res` below is the significant subset, and the
# non-significant rows are the background pool for slSampleNonEventExons()
dtu_res <- rowData(se)[["fitDTUResult_WT_vs_KD"]] |>
  as.data.frame() |>
  tibble::rownames_to_column("tx_name") |>
  tibble::as_tibble() |>
  dplyr::mutate(gene_id = gene_ids) |>
  dplyr::select(tx_name, gene_id, estimates, pval, regular_FDR) |>
  dplyr::arrange(pval)

sig_res <- dtu_res |> dplyr::filter(regular_FDR < 0.3)

sig_res <- sig_res |> mutate( tx_id = tx_name)
sig_res
# save sig_res for later use
saveRDS(sig_res, here("workflows/data/sig_res.rds"))
```

```{r load_sig_res}
# load the saved DTU results
sig_res <- readRDS(here("workflows/data/sig_res.rds"))
sig_res
```

```{r txdb, eval=FALSE}
library(txdbmaker)
# load TxDb for GRanges conversion
# download once (warning: ~50MB compressed)
wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_48/gencode.v48.annotation.gtf.gz

txdb <- txdbmaker::makeTxDbFromGFF(
  "/Users/beacm/Documents/splicelogic-paper/workflows/gencode.v48.annotation.gtf.gz",
  format = "gtf"
)
# save to not rebuild every time
AnnotationDbi::saveDb(txdb, here("workflows/data/gencode.v48.sqlite"))

# check how many transcripts in se are in txdb
rowData(se)$tx_name %in% AnnotationDbi::keys(txdb, "TXNAME") |> sum() / length(rowData(se)$tx_name)
```
## Splice event detection with splicelogic

```{r make_exons, eval=FALSE}
# build exon GRangesList and rename from internal TXID to TXNAME
txdb <- AnnotationDbi::loadDb(here("workflows/data/gencode.v48.sqlite"))
exons <- exonsBy(txdb, by = "tx")
txps <- AnnotationDbi::select(txdb, keys(txdb, "TXID"), c("TXNAME", "GENEID"), "TXID") |>
  tibble::as_tibble() |>
  dplyr::mutate(TXID = as.character(TXID))
# rename exons to match tx_name in se, matching by TXID
all.equal(names(exons), txps$TXID)
names(exons) <- txps$TXNAME[match(names(exons), txps$TXID)]
saveRDS(exons, here("workflows/data/glinos_exons.rds"))
```

```{r input_formatting}
# load from saved sqlite
txdb <- AnnotationDbi::loadDb(here("workflows/data/gencode.v48.sqlite"))
# load exons from saved rds
exons <- readRDS(here("workflows/data/glinos_exons.rds"))
```

```{r preprare_exons}
# prepare exons for splicelogic
sig_exons <-  prepare_exons(
  txdb, sig_res, coef_col = "estimates", verbose = TRUE
)
```

```{r find_events}
#find se
se_events <- sig_exons |> preprocess(coef_col = "estimates")|> find_se()

#find all events
events <- sig_exons |> preprocess(coef_col = "estimates")|> find_all_events()
```


```{r collapse_redundant_events}
# define a helper function
my_paste <- \(x) paste(unique(x), collapse = ",")
collapsed_events <- events |>
    dplyr::group_by(exon_id, event_type) |>
    plyranges::reduce_ranges_directed(
        n = plyranges::n(),
        tx_id = my_paste(tx_id),
        exon_rank = my_paste(exon_rank),
        event_tx_id = my_paste(event_tx_id),
  )
```

```{r barplot_event_types}
event_counts <- events |>
    as_tibble() |>
    dplyr::group_by(event_type) |>
    dplyr::summarise(count = dplyr::n())

collapsed_event_counts <- collapsed_events |>
    as_tibble() |>
    dplyr::group_by(event_type) |>
    dplyr::summarise(count = dplyr::n())

# combine both count tables into one, labelled by source
counts <- dplyr::bind_rows(
  "All events" = event_counts,
  "Collapsed events" = collapsed_event_counts,
  .id = "set"
)

ggplot(counts, aes(x = event_type, y = count, fill = set)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c(
    "All events" = "#9ecae1",       # light blue
    "Collapsed events" = "#3182bd"  # darker blue
  )) +
  labs(title = "Splice event counts", 
  x = "Event Type", y = "Count", fill = NULL) +
  theme_bw()
```

```{r plot_events_by_gene}
# use splicelogic's built-in plotting function to visualize events for a specific gene
# for example, let's visualize events for the gene "ENSG00000187634.13"
gene_of_interest <- "ENSG00000196924.19"
# filter events for the gene of interest
events_for_gene <- events[events$gene_id == gene_of_interest]

# visualize the events for the gene of interest
splicelogic::plot_transcripts(events_for_gene, sig_exons, rescale_introns = FALSE) +
  ggtitle(paste("Splice events for gene:", gene_of_interest))
 
```

```{r save_events_plots, eval=FALSE}
# make a plot for each gene with events and save to files
# plot_transcripts_by_gene returns a list of ggplot objects, one for each gene
plots <- plot_transcripts_by_gene(events, sig_exons)
# save each plot to the same pdf file - takes a while: -- 1725 genes 
pdf(here("workflows/figures/splice_events_by_gene.pdf"), width = 8, height = 6)
for (gene in names(plots)) {
    print(plots[[gene]])
}
dev.off()
```

## Sequence context around the event exons, focusing on skipped exons (SE) and included exons (IE)

The analysis below uses a set of sequence-context helpers consolidated in
`helper.R` (nucleotide-composition profiles, background exon sampling,
splice-site logos and RBP motif scanning). The splicelogic package itself only
exports `get_seq()` for sequence work, so these stay as paper-repo helpers (for now).

```{r seq_context_helpers, message=FALSE}
source(here("workflows/helper.R"))
```

`slEventFlankProfiles()` takes a finder's output straight away: it flanks each
exon (strand-aware, so "upstream" is 5' in transcription order), slides
overlapping windows across the flank, and returns the per-window nucleotide
fractions for each side.

```{r flank_windows_events, message=FALSE}
se_profiles <- slEventFlankProfiles(se_events, genome = "hg38", width = 100)

# rows = events, cols = window_nucleotide (e.g. w1_A)
dim(se_profiles$upstream)
dim(se_profiles$downstream)
```

```{r plot_updownstream_events, fig.width=8, fig.height=3.5}
slPlotFlankProfiles(se_profiles, exon_label = "SE")
```

### included exons events

```{r flank_windows_ie, message=FALSE}

ie_events <- sig_exons |> preprocess(coef_col = "estimates")|> find_ie()

ie_profiles <- slEventFlankProfiles(ie_events, genome = "hg38", width = 100)

# rows = events, cols = window_nucleotide (e.g. w1_A)
dim(ie_profiles$upstream)
dim(ie_profiles$downstream)
```

```{r plot_updownstream_ie, fig.width=8, fig.height=3.5}
slPlotFlankProfiles(ie_profiles, exon_label = "IE")
```


## Background set

We want to compare the sequence context of the events with that of a control set.
For the control set (non-event exons), we sample as many exons as there are events, 
drawn at random from the internal exons of transcripts with no evidence of 
differential usage (`pval > 0.5`).

```{r non_event_exons}
non_event <- slSampleNonEventExons(
  dtu_res, exons, n = se_events, pval_min = 0.5, seed = 1
)
non_event
```

```{r flank_windows_non_event, message=FALSE, fig.width=8, fig.height=3.5}
non_event_profiles <- slEventFlankProfiles(
  non_event, genome = "hg38", width = 100
)
dim(non_event_profiles$upstream)

slPlotFlankProfiles(non_event_profiles, exon_label = "non-SE")
```

An alternative, within-transcript control: for each event, the internal exon of
the same transcript that is shared with the event partner and closest in rank
to the event exon.

```{r matched_non_event_exons, eval=FALSE, fig.width=8, fig.height=3.5}
matched <- slMatchedNonEventExons(se_events, exons)

slEventFlankProfiles(matched, genome = "hg38", width = 100) |>
  slPlotFlankProfiles(exon_label = "matched exon")
```

## Splice-site sequence logos

The window profiles above average over 10 bp, which smooths the splice site
away. `slFlankLogoMatrices()` keeps every base position separate and returns a
position frequency matrix per side, running `width` bases into the intron and
`exon_bases` into the exon so the boundary is inside the matrix.

Note which site each side carries: the region **upstream** of an exon is the
end of the preceding intron and holds the **3' splice site** (acceptor,
`...ag | EXON`); the region **downstream** is the start of the following intron
and holds the **5' splice site** (donor, `EXON | gu...`).

```{r logo_events, message=FALSE, warning=FALSE, fig.width=13, fig.height=3.2}
se_logo <- slFlankLogoMatrices(se_events, genome = "hg38",
                               width = 25, exon_bases = 3)

# letter height = frequency at that position
slPlotFlankLogo(se_logo, method = "prob")
```

```{r logo_events_bits, message=FALSE, warning=FALSE, fig.width=13, fig.height=3.2}
# letter height = information content, the classic logo scaling: unconstrained
# positions shrink to nothing and the consensus stands out
slPlotFlankLogo(se_logo, method = "bits")
```

The same for the background set, to see whether the event splice sites differ
from those of exons in transcripts with no differential usage.

```{r logo_non_event, message=FALSE, warning=FALSE, fig.width=13, fig.height=3.2}
non_event_logo <- slFlankLogoMatrices(non_event, genome = "hg38",
                                      width = 25, exon_bases = 3)

slPlotFlankLogo(non_event_logo, method = "prob")
```

```{r logo_difference}
# where the two sets differ most, per position and nucleotide
delta <- se_logo$downstream - non_event_logo$downstream
round(delta[, order(-apply(abs(delta), 2, max))[1:6]], 3)
```
